### 2019-09-11 reports/thesis/chapters/3_compoundPoisson.tex
Revise

The compound Poisson has applications in, for example, modelling rainfall \citep{revfeim1984initial} and insurance claims \citep{jorgensen1994fitting, smyth2002fitting}.

This chapter starts with a literature review on the compound Poisson distribution, how it is derived from the behaviour of photons, how its likelihood is evaluated and methods for fitting it onto data. A model was proposed for the pixel's grey values and the expectation-maximization (EM) algorithm was implemented to fit the model onto data. It was found that for high photon rate, there were identifiability issues. The chapter is concluded on a discussion on why the EM algorithm failed.

@@

In an x-ray tube, photons are emitted as a Poisson process \citep{whiting2006properties, cierniak2011x} and each photon has some random energy due to bremsstrahlung and characteristic radiation \citep{sun2012overview}. This shares similarities to the compound Poisson. Let $Y$ be the number of photons emitted for some time exposure $\tau$, then $Y\sim\poisson(\lambda)$. Each photon is assumed to be i.i.d.~with energy $U_i$ for $i=1,2,3,\dotsc$. $U_i$ has p.d.f.~$p_U(u)$. The random variables discussed here covers all the latent variables in the compound Poisson.

Photons emitted from the x-ray tube undergo attenuation when propagating through the test sample. Assuming no beam hardening, some photons are either absorbed or scattered, making them undetectable. Scattered photons may be detected but it is very rare \citep{cantatore2011introduction}. Attenuated photon energy remains unaffected so attenuation decreases the parameter $\lambda$ and the amount it decreases by depends on the attenuation coefficient of the material and the amount of material the x-ray attenuates. The parameters of the random variable $U_i$ remains unchanged because the energy of each photon remains the same after attenuation, assuming no beam hardening.

When the photons interact with the scintillator in the detector, they are converted into visible light. The visible light photons are then detected and converted into a digital signal. A quantum counter set the digital signal to be linear with to the number of photons detected \citep{whiting2006properties}. Let $X$ be the grey value observed, then
\begin{equation}
X = bY + \epsilon
\end{equation}
where $\epsilon\sim\normal(a,\kappa)$, $b$ and $a$ are some constant and $\kappa$ is the variance of electronic noise. The mean and variance of the grey value are
\begin{equation}
\expectation\left[X\right] = b\lambda + a
\end{equation}
and
\begin{equation}
\variance\left[X\right] = b^2\lambda + \kappa
\end{equation}
respectively. By eliminating $\lambda$
\begin{equation}
\variance\left[X\right] = b\expectation\left[X\right]+\kappa-ab \ ,
\end{equation}
a linear relationship between the variance and expectation of the grey value \citep{ma2012varaince} is obtained.

An energy integrating detector records the grey value as linear to the energy detected \citep{whiting2006properties}. The grey value $X$ is
\begin{equation}
X|Y = \sum_{i=1}^Y U_i + \epsilon \ .
\end{equation}
This is the compound Poisson with Normal noise added to it. The scale factor $b$ is not included as this can be absorbed into $U$. Using the result that $\expectation\left[X\right]=\expectation\expectation\left[X|Y\right]$ and $\variance\left[X\right] = \variance\expectation\left[X|Y\right] + \expectation\variance\left[X|Y\right]$, the mean and variance of the grey value are
\begin{equation}
\expectation\left[X\right] = \lambda\expectation\left[U\right]+a
\end{equation}
and
\begin{equation}
\variance\left[X\right] = \lambda \expectation\left[U^2\right]+\kappa
\end{equation}
respectively. Eliminating $\lambda$ obtains
\begin{equation}
\variance\left[X\right] = \dfrac{\expectation\left[U^2\right]}{\expectation\left[U\right]} \expectation\left[X\right] + \kappa - a\dfrac{\expectation\left[U^2\right]}{\expectation\left[U\right]} \ .
\end{equation}
By assuming no beam hardening, $\expectation\left[U\right]$ and $\expectation\left[U^2\right]$ remains constant, thus there is a linear relationship between the variance and expectation of the grey value \citep{yang2009evaluation}. There are other types of detection schemes \citep{whiting2006properties} but it shall be not be considered here.

Experiments have been done to verify the compound Poisson nature of the detector. This was done by investigating the variance of radiographs of air \citep{hsieh2015compound} and a polyethylene cylinder \citep{yang2009evaluation, yang2010noise} at different voltages and powers. It was found there were two components in the noise, one was signal dependent and comes from the compound Poisson, the other was signal independent and may be electronic noise. The electronic noise can be modelled using a Normal distribution \citep{xu2009electronic}.

@@

Returning back to the compound Poisson with no electronic noise $X|Y = \sum_{i=1}^{Y}U_i$. Let the moment generating function (m.g.f.) of $X$ be $M_X(\theta)=\expectation\left[\euler^{X\theta}\right]$. It can be shown that the m.g.f.~is
\begin{equation}
  M_X(\theta)=
  \exp\left[
    \lambda
    \left(
      M_U(\theta)-1
    \right)
  \right]
\end{equation}
\citep{gatto2010saddlepoint}. The derivation is shown in Appendix \ref{chapter:appendix_compoundPoissonMgf}. Moments of $X$ can be obtained from the m.g.f.~by differentiating it and setting $\theta$ to zero. In other words, $\expectation[X^r]=M_X^{(r)}(0)$. Then it can be shown that
\begin{align}
  \expectation\left[X\right]&=\lambda\expectation\left[U\right]
  \\
  \variance\left[X\right] &= \lambda\expectation\left[U^2\right]
  \\
  \expectation\left[(X-\expectation[X])^3\right] &=\lambda\expectation\left[U^3\right]
\end{align}
which agrees with the expectation and variance results in the previous section. The derivation is shown in Appendix \ref{chapter:appendix_compoundPoissonMgf}.

\subsection{Compound Poisson-Gamma}
\label{chapter:compoundPoisson_compoundPoissonGamma}
A special case of the compound Poisson is when $U\sim\gammaDist\left(\alpha,\beta\right)$ where $\alpha>0$ is the gamma shape parameter and $\beta>0$ is the gamma rate parameter. This was used for example in \cite{xu2009electronic}. This is known as the compound Poisson-gamma distribution. This is denoted by $X\sim\CPoisson(\lambda,\alpha,\beta)$ and has p.d.f.
\begin{equation}
  p_X(x) = 
  \begin{cases}
    \delta(x) \euler^{-\lambda} & \text{ for } x=0 \\ 
    \displaystyle\sum_{y=1}^{\infty}\dfrac{\beta^{y\alpha}}{\Gamma(y\alpha)}x^{y\alpha-1}\euler^{-\beta x}\euler^{-\lambda}\frac{\lambda^y}{y!} & \text{ for } x>0
  \end{cases}
  \ .
  \label{eq:compoundPoisson_pdf}
\end{equation}
Recall that $X|Y=\sum_{i=1}^YU_i$ which involves a sum of gamma random variables. It can be shown that
\begin{equation}
  X|Y\sim\gammaDist\left(Y\alpha,\beta\right).
\end{equation}

@@

It can be shown that the compound Poisson-gamma distribution is in the exponential family for fixed $\alpha$ \citep{jorgensen1987exponential}. To show this, it is useful to parametrise the compound Poisson-gamma distribution using the following:
\begin{equation}
  p=\frac{2+\alpha}{1+\alpha}
  \ ,
\end{equation}
\begin{equation}
  \mu=\frac{\lambda\alpha}{\beta}
  \ ,
\end{equation}
\begin{equation}
  \phi = \frac{\alpha+1}{\beta^{2-p}(\lambda\alpha)^{p-1}}
  \ .
\end{equation}
The parameters $p$, $\mu$ and $\phi$ are called the index, mean and dispersion parameters respectively. The parameters take the values of $1<p<2$, $\mu>0$ and $\phi>0$. After parametrising, it can be shown that the p.m.f.~at zero is
\begin{equation}
  \prob(X=0) = \exp
  \left[
      -\frac{\mu^{2-p}}{\phi(2-p)}
  \right]
\end{equation}
and the p.d.f.~for $x>0$ is
\begin{equation}
  p_X(x) = 
  \exp\left[
    \frac{1}{\phi}
    \left(
      x\frac{\mu^{1-p}}{1-p}-\frac{\mu^{2-p}}{2-p}
    \right)
  \right]
  \frac{1}{x}
  \sum_{y=1}^{\infty}W_y(x,p,\phi)
\end{equation}
where
\begin{equation}
  W_y = W_y(x,p,\phi)=\frac{x^{y\alpha}}{\phi^{y(1+\alpha)}(p-1)^{y\alpha}(2-p)^yy!\Gamma(y\alpha)}
  \ .
\end{equation}
The derivation is shown in Appendix \ref{chapter:appendix_tweedie}. This is in the form of a generalised linear model \citep{nelder1972generalized, nelder1972generalized_2, mccullagh1984generalized} for fixed $p$ and $\phi$ because the above is in the form of a distribution in the dispersive exponential family.

Parameter estimation for fixed $p$ can be done via the generalised linear model framework and can be extended to include linear mixed models \citep{zhang2013likelihood}. Estimating $p$ is difficult and various methods were discussed by \cite{zhang2013likelihood}. One way is to estimate $\mu$ and $\phi$ on a grid of $p$'s and then select the $p$ which maximises the likelihood \citep{dunn2005series}.

@@

The method of moments is a simpler method to estimate the parameters of a compound Poisson-gamma random variable. Suppose $\widehat{\mu}_1$ is an estimator of $\expectation[X]$ and $\widehat{\mu}_j$ is an estimator of $\expectation\left[\left(X-\expectation[X]\right)^j\right]$ for $j=2,3$. Then the estimators
\begin{equation}
  \widehat{\lambda}=\frac{\widehat{\mu}_1^2\widehat{\mu}_2}{2\widehat{\mu}_2^2-\widehat{\mu}_1\widehat{\mu}_3}
\end{equation}
\begin{equation}
  \widehat{\alpha}=\frac{2\widehat{\mu}_2^2-\widehat{\mu}_1\widehat{\mu}_3}{\widehat{\mu}_1\widehat{\mu}_3-\widehat{\mu}_2^2}
\end{equation}
\begin{equation}
  \widehat{\beta}=\frac{\widehat{\mu}_1\widehat{\mu}_2}{\widehat{\mu}_1\widehat{\mu}_3-\widehat{\mu}_2^2}
\end{equation}
are method of moments estimators of $\lambda$, $\alpha$ and $\beta$ respectively \citep{withers2011compound}, this is shown in Appendix \ref{chapter:appendix_methodofmoments} These estimators suffer because estimation is not done through the sufficient statistics and can be negative, this is a problem because the parameters do not take non-positive values.

@@

The evaluation of the density of a compound Poisson-gamma distribution is useful so that the likelihood can be obtained. The likelihood then can be used to find, for example, maximum likelihood estimators. A problem occurs when dealing with the infinite sum in the p.d.f.~because it cannot be simplified. There are a number of approximations or computational methods to evaluate the p.d.f.~such as Fourier inverting the characteristic function \citep{dunn2008evaluation}, using the saddlepoint approximation \citep{daniels1954saddlepoint} or cleverly sum over certain terms in the infinite sum \citep{dunn2005series}. Monte Carlo methods can be used to evaluate the p.d.f.~by simulating compound Poisson-gamma random variables.

@@

The saddlepoint approximation \citep{daniels1954saddlepoint} gives an approximate solution to inverting the Laplace transformation or the m.g.f.~giving the p.d.f.~and has been used in XCT \citep{elbakri2003statistical, elbakri2003efficient}. Inverting the Fourier transformation of the characteristic function also gives the p.d.f.~using computational methods \citep{dunn2008evaluation}. The saddlepoint approximation will be studied here.

For a given m.g.f.~$M_X(\theta)$, the saddlepoint approximation \citep{daniels1954saddlepoint, butler2007saddlepoint} finds an approximate p.d.f.~$p_X(x)$. The saddlepoint approximation is given as
\begin{equation}
  p_X(x)\approx\left(2\pi K_X''(s)\right)^{-1/2}\exp\left[K_X(s)-sx\right]
  \label{eq:saddlePoint:generalSaddlePoint}
\end{equation}
where $K_X(\theta) = \ln\left(M_X(\theta)\right)$, and $s=s(x)$ is the solution to the saddle point equation $K_X'(s)=x$.

@@

The integral of the density approximation over the support may not equal to one, it can be numerically re-normalised if necessary. Thus it may be more sensible to write the approximation up to a constant
\begin{equation}
  p_X(x)\propto x^{-\frac{\alpha+2}{2(\alpha+1)}}\euler^{-x\beta}
  \exp\left[
    x^{\frac{\alpha}{\alpha+1}}
    \frac{
      (\lambda\beta^\alpha)^{\frac{1}{\alpha+1}}(\alpha+1)
    }
    {
      \alpha^{\frac{\alpha}{\alpha+1}}
    }
  \right]
  \ .
\end{equation}

@@

The infinite sum, $\sum_{y=1}^\infty W_y$, can be computationally summed in a clever way in order to evaluate the p.d.f. This was done by summing only large terms in the sum and ignore small terms \citep{dunn2005series}. \cite{dunn2005series} approximated the sum by truncation
\begin{equation}
  \sum_{y=1}^\infty W_y \approx \sum_{y=y_\text{l}}^{y_\text{u}}W_y
\end{equation}
where $y_\text{l}<y_{\text{max}}<y_\text{u}$ and $y_{\text{max}}$ is the value of $y$ which maximises $W_y$. \cite{dunn2005series} used Stirling's approximation to find that
\begin{equation}
  y_{\text{max}} \approx \frac{x^{2-p}}{\phi(2-p)}
\end{equation}
by treating $W_y$ as a continuous and differentiable function of $y$. The derivation is shown in Appendix \ref{chapter:appendix_compoundPoissonSeries}. Because values of $y$ should be a positive integer, it would be appropriate to round the solution to $y_\text{max}$ accordingly
\begin{equation}
  y_{\text{max}} = \text{max}\left[
    1,\text{round}\left(\frac{x^{2-p}}{\phi(2-p)}\right)
  \right]
  \ .
\end{equation}

$y_\text{l}$ and $y_\text{u}$ can be chosen such that $W_{y_\text{l}}$ and $W_{y_\text{u}}$ are less than $\epsilon W_{y_\text{max}}$ where $\epsilon$ is some small constant, for example $\epsilon=\euler^{-37}$ will be better than machine precision in 64 bits \citep{dunn2005series}. To prevent overflow problems, it is advised to calculate each term in the summation in log scale \citep{dunn2005series} by using the following equation
\begin{equation}
  \ln\left[
    \sum_{y=y_\text{l}}^{y_\text{u}}W_y
  \right]
  = 
  \ln\left(
    W_{y_\text{max}}
  \right)
  +\ln\sum_{y=y_\text{l}}^{y_\text{u}}
  \exp\left[
    \ln\left(W_y\right)-\ln\left(W_{y_\text{max}}\right)
  \right]
  \ .
\end{equation}

@@

Simulations of a compound Poisson-gamma random variable were conducted with the aim to compare how well these density evaluations methods perform. This was done by comparing the evaluated densities with the histogram of simulations and a Q-Q plot. The following compound Poisson-gamma distributions were used in the simulations to capture the variety in the compound Poisson-gamma family: $\CPoisson(1,1,1)$, $\CPoisson(1,100,1)$, $\CPoisson(10,1,1)$, $\CPoisson(100,100,1)$. Varying $\beta$ is not interesting here as this only scales the random variable.

There were a few technical problems with the histogram because the compound Poisson has probability mass at zero and probability density for positive numbers. To correctly represent the empirical distribution of a compound Poisson random variable, a bar chart was used to show the frequency of a zero and a histogram to show the frequency density of positive numbers. However, the Normal approximation and the saddlepoint approximation does not have mass support at zero. Therefore, to compare these approximate densities to the empirical distribution fairly, a histogram containing both zero and positive samples was used when appropriate.

The evaluation of the p.d.f.~using the saddlepoint approximate required a bit of caution to avoid over/underflow problems. Suppose realizations of $X$ were simulated $\left\{x_1,x_2,x_3,\dotsc,x_n\right\}$. The saddlepoint approximation was computed up to a constant using
\begin{equation}
  p_X(x) \propto
  \exp\left[
    -\frac{\alpha+2}{2(\alpha+1)}
    \ln(x)
    -x\beta
    +\left(
    \frac{x\beta}{\alpha}
    \right)^{\frac{\alpha}{\alpha+1}}\lambda^{\frac{1}{\alpha+1}}(\alpha+1) - k
  \right]
\end{equation}
for 10\,000 equally spaced points from and including the minimum non-zero simulated value to the maximum simulated value. $k$ is some constant which was chosen to be
\begin{equation}
  k =
  \max_{i\in\left\{1,2,3,\dotsc,n\right\}}
  \left[
    -\frac{\alpha+2}{2(\alpha+1)}
    \ln(x_i)
    -x_i\beta
    +\left(
      \frac{x_i\beta}{\alpha}
    \right)^{\frac{\alpha}{\alpha+1}}\lambda^{\frac{1}{\alpha+1}}(\alpha+1)
  \right]
  \ .
\end{equation}
The density was then was normalised by numerically integrating it using the trapezium rule using the 10\,000 evaluated points.

>>>>

The compound Poisson distribution has applications in, for example, modelling rainfall \citep{revfeim1984initial} and insurance claims \citep{jorgensen1994fitting, smyth2002fitting}.

This chapter starts with a literature review on the compound Poisson distribution, how it is derived from the behaviour of photons, how its likelihood is evaluated and methods for fitting it onto data. A model was proposed for the grey values and the expectation-maximisation (EM) algorithm was implemented to fit the model onto data. It was found that for high photon rates, there were identifiability issues. The chapter is concluded on a discussion on why the EM algorithm failed.

@@

In an x-ray tube, photons are emitted as a Poisson process \citep{whiting2006properties, cierniak2011x} and each photon has some random energy due to bremsstrahlung and characteristic radiation \citep{sun2012overview}. This shares similarities to the compound Poisson distribution. Let $Y$ be the number of photons emitted for some time exposure $\tau$, then $Y\sim\poisson(\lambda)$. Each photon is assumed to be i.i.d.~with random energy $U_i$ for $i=1,2,3,\dotsc$ with p.d.f.~$p_U(u)$. The random variables discussed here cover all the latent variables in the compound Poisson.

Photons emitted from the x-ray tube undergo attenuation when propagating through the test sample. Assuming no beam hardening, some photons are either absorbed or scattered, making them undetectable. Scattered photons may be detected but it is very rare \citep{cantatore2011introduction}. The energy of the attenuated photons remain unaffected so attenuation decreases the parameter $\lambda$. The amount it decreases by depends on the attenuation coefficient of the material and the amount of material the x-ray attenuates. The density of the random variable $U_i$ remains unchanged because the energy of each photon remains the same after attenuation, assuming no beam hardening.

When the photons interact with the scintillator in the detector, they are converted into visible light. The visible light photons are then detected and converted into a digital signal or grey value. A quantum counter set the digital signal to be linear with the number of photons detected \citep{whiting2006properties}. Let $X$ be the grey value observed, then
\begin{equation}
X = bY + \epsilon
\end{equation}
where $\epsilon\sim\normal(a,\kappa)$, $b$ and $a$ are some constant and $\kappa$ is the variance of electronic noise. The mean and variance of the grey value are
\begin{equation}
\expectation\left[X\right] = b\lambda + a
\end{equation}
and
\begin{equation}
\variance\left[X\right] = b^2\lambda + \kappa
\end{equation}
respectively. By eliminating $\lambda$
\begin{equation}
\variance\left[X\right] = b\expectation\left[X\right]+\kappa-ab \ ,
\end{equation}
a linear relationship between the variance and expectation of the grey value \citep{ma2012varaince} is obtained.

In an energy integrating detector, the recorded grey value is linear to the energy detected \citep{whiting2006properties}. The grey value $X$ is
\begin{equation}
X|Y = \sum_{i=1}^Y U_i + \epsilon \ .
\end{equation}
This is the compound Poisson distribution with Normal noise added to it. The scale factor $b$ is not included as this can be absorbed into $U$. Using the result that $\expectation\left[X\right]=\expectation\expectation\left[X|Y\right]$ and $\variance\left[X\right] = \variance\expectation\left[X|Y\right] + \expectation\variance\left[X|Y\right]$, the mean and variance of the grey value are
\begin{equation}
\expectation\left[X\right] = \lambda\expectation\left[U\right]+a
\end{equation}
and
\begin{equation}
\variance\left[X\right] = \lambda \expectation\left[U^2\right]+\kappa
\end{equation}
respectively. Eliminating $\lambda$ obtains
\begin{equation}
\variance\left[X\right] = \dfrac{\expectation\left[U^2\right]}{\expectation\left[U\right]} \expectation\left[X\right] + \kappa - a\dfrac{\expectation\left[U^2\right]}{\expectation\left[U\right]} \ .
\end{equation}
By assuming no beam hardening, $\expectation\left[U\right]$ and $\expectation\left[U^2\right]$ remains constant, thus there is a linear relationship between the variance and expectation of the grey value \citep{yang2009evaluation}. There are other types of detection schemes \citep{whiting2006properties} but it shall be not be considered here.

Experiments have been done to verify the compound Poisson nature of the detector. This was done by investigating the variance of radiographs of air \citep{hsieh2015compound} and a polyethylene cylinder \citep{yang2009evaluation, yang2010noise} using different exposures. It was found there were two components in the noise, one was signal dependent and comes from the compound Poisson distribution, the other was signal independent and may be electronic noise. The electronic noise can be modelled using a Normal distribution \citep{xu2009electronic}.

@@

Returning back to the compound Poisson with no electronic noise $X|Y = \sum_{i=1}^{Y}U_i$, let the moment generating function (m.g.f.) of $X$ be $M_X(\theta)=\expectation\left[\euler^{X\theta}\right]$. It can be shown that the m.g.f.~is
\begin{equation}
  M_X(\theta)=
  \exp\left[
    \lambda
    \left(
      M_U(\theta)-1
    \right)
  \right]
\end{equation}
\citep{gatto2010saddlepoint}. The derivation is shown in Appendix \ref{chapter:appendix_compoundPoissonMgf}. Moments of $X$ can be obtained from the m.g.f.~by differentiating it and setting $\theta$ to zero. In other words, $\expectation[X^r]=M_X^{(r)}(0)$. Then it can be shown that
\begin{align}
  \expectation\left[X\right]&=\lambda\expectation\left[U\right]
  \\
  \variance\left[X\right] &= \lambda\expectation\left[U^2\right]
  \\
  \expectation\left[(X-\expectation[X])^3\right] &=\lambda\expectation\left[U^3\right]
\end{align}
which agrees with the expectation and variance results in the previous section. The derivation is shown in Appendix \ref{chapter:appendix_compoundPoissonMgf}.

\subsection{Compound Poisson-Gamma Distribution}
\label{chapter:compoundPoisson_compoundPoissonGamma}
A special case of the compound Poisson is when $U\sim\gammaDist\left(\alpha,\beta\right)$ where $\alpha>0$ is the gamma shape parameter and $\beta>0$ is the gamma rate parameter. This was used for example in \cite{xu2009electronic}. The distribution is known as the compound Poisson-gamma distribution and is denoted by $X\sim\CPoisson(\lambda,\alpha,\beta)$. The p.d.f.~is
\begin{equation}
  p_X(x) = 
  \begin{cases}
    \delta(x) \euler^{-\lambda} & \text{ for } x=0 \\ 
    \displaystyle\sum_{y=1}^{\infty}\dfrac{\beta^{y\alpha}}{\Gamma(y\alpha)}x^{y\alpha-1}\euler^{-\beta x}\euler^{-\lambda}\frac{\lambda^y}{y!} & \text{ for } x>0
  \end{cases}
  \ .
  \label{eq:compoundPoisson_pdf}
\end{equation}
Recall that $X|Y=\sum_{i=1}^YU_i$ which involves a sum of gamma random variables. It can be shown that
\begin{equation}
  X|Y\sim\gammaDist\left(Y\alpha,\beta\right).
\end{equation}

@@

It can be shown that the compound Poisson-gamma distribution is in the exponential family for fixed $\alpha$ \citep{jorgensen1987exponential}. To show this, the compound Poisson-gamma distribution was parametrised using the following:
\begin{equation}
  p=\frac{2+\alpha}{1+\alpha}
  \ ,
\end{equation}
\begin{equation}
  \mu=\frac{\lambda\alpha}{\beta}
  \ ,
\end{equation}
\begin{equation}
  \phi = \frac{\alpha+1}{\beta^{2-p}(\lambda\alpha)^{p-1}}
  \ .
\end{equation}
The parameters $p$, $\mu$ and $\phi$ are called the index, mean and dispersion parameters respectively and take the values of $1<p<2$, $\mu>0$ and $\phi>0$. It can be shown that the p.m.f.~at zero is
\begin{equation}
  \prob(X=0) = \exp
  \left[
      -\frac{\mu^{2-p}}{\phi(2-p)}
  \right]
\end{equation}
and the p.d.f.~for $x>0$ is
\begin{equation}
  p_X(x) = 
  \exp\left[
    \frac{1}{\phi}
    \left(
      x\frac{\mu^{1-p}}{1-p}-\frac{\mu^{2-p}}{2-p}
    \right)
  \right]
  \frac{1}{x}
  \sum_{y=1}^{\infty}W_y(x,p,\phi)
\end{equation}
where
\begin{equation}
  W_y = W_y(x,p,\phi)=\frac{x^{y\alpha}}{\phi^{y(1+\alpha)}(p-1)^{y\alpha}(2-p)^yy!\Gamma(y\alpha)}
  \ .
\end{equation}
The derivation is shown in Appendix \ref{chapter:appendix_tweedie}. This is in the form of a distribution in the dispersive exponential family.\citep{nelder1972generalized, nelder1972generalized_2, mccullagh1984generalized} for fixed $p$ and $\phi$.

Parameter estimation for known $p$ can be done via the generalised linear model framework and can be extended to include linear mixed models \citep{zhang2013likelihood}. Estimating $p$ is difficult and various methods were discussed \citep{zhang2013likelihood}. One way is to estimate $\mu$ and $\phi$ on a grid of $p$'s and then select the $p$ which maximises the likelihood \citep{dunn2005series}.

@@

The method of moments is a simpler method to estimate the parameters of a compound Poisson-gamma distribution. Suppose $\widehat{\mu}_1$ is an estimator of $\expectation[X]$ and $\widehat{\mu}_j$ is an estimator of $\expectation\left[\left(X-\expectation[X]\right)^j\right]$ for $j=2,3$, then the estimators
\begin{equation}
  \widehat{\lambda}=\frac{\widehat{\mu}_1^2\widehat{\mu}_2}{2\widehat{\mu}_2^2-\widehat{\mu}_1\widehat{\mu}_3}
\end{equation}
\begin{equation}
  \widehat{\alpha}=\frac{2\widehat{\mu}_2^2-\widehat{\mu}_1\widehat{\mu}_3}{\widehat{\mu}_1\widehat{\mu}_3-\widehat{\mu}_2^2}
\end{equation}
\begin{equation}
  \widehat{\beta}=\frac{\widehat{\mu}_1\widehat{\mu}_2}{\widehat{\mu}_1\widehat{\mu}_3-\widehat{\mu}_2^2}
\end{equation}
are method of moments estimators of $\lambda$, $\alpha$ and $\beta$ respectively \citep{withers2011compound}, this is shown in Appendix \ref{chapter:appendix_methodofmoments}. These estimators suffer because estimation is not done through the sufficient statistics and can be negative. This is a problem because the parameters do not take non-positive values.

@@

The evaluation of the p.d.f.~of a compound Poisson-gamma distribution is useful so that the likelihood can be obtained. The likelihood can be used to find, for example, maximum likelihood estimators. A problem occurs when dealing with the infinite sum in the p.d.f.~because it cannot be simplified. There are a number of approximations or computational methods to evaluate the p.d.f.~such as Fourier inverting the characteristic function \citep{dunn2008evaluation}, using the saddlepoint approximation \citep{daniels1954saddlepoint} or cleverly sum over certain terms in the infinite sum \citep{dunn2005series}. Monte Carlo methods can be used to evaluate the p.d.f.~by simulating compound Poisson-gamma random variables.

@@

The saddlepoint approximation \citep{daniels1954saddlepoint} uses the approximate solution to inverting the Laplace transformation or the m.g.f. This has been used in XCT \citep{elbakri2003statistical, elbakri2003efficient}. Inverting the Fourier transformation of the characteristic function also gives the p.d.f.~using computational methods \citep{dunn2008evaluation}. The saddlepoint approximation will be studied here.

For a given m.g.f.~$M_X(\theta)$, the saddlepoint approximation \citep{daniels1954saddlepoint, butler2007saddlepoint} finds an approximate p.d.f.~$p_X(x)$. The saddlepoint approximation is given as
\begin{equation}
  p_X(x)\approx\left(2\pi K_X''(s)\right)^{-1/2}\exp\left[K_X(s)-sx\right]
  \label{eq:saddlePoint:generalSaddlePoint}
\end{equation}
where $K_X(\theta) = \ln\left(M_X(\theta)\right)$ and $s=s(x)$ is the solution to the saddle point equation $K_X'(s)=x$.

@@

The integral of the density approximation over the support may not equal to one and it can be numerically re-normalised if necessary. Thus, it may be more sensible to write the approximation up to a constant
\begin{equation}
  p_X(x)\propto x^{-\frac{\alpha+2}{2(\alpha+1)}}\euler^{-x\beta}
  \exp\left[
    x^{\frac{\alpha}{\alpha+1}}
    \frac{
      (\lambda\beta^\alpha)^{\frac{1}{\alpha+1}}(\alpha+1)
    }
    {
      \alpha^{\frac{\alpha}{\alpha+1}}
    }
  \right]
  \ .
\end{equation}

@@

The infinite sum, $\sum_{y=1}^\infty W_y$, can be computationally summed in a clever way in order to evaluate the p.d.f. This was done by summing only large terms in the sum and ignoring small terms \citep{dunn2005series}. \cite{dunn2005series} approximated the sum by truncation
\begin{equation}
  \sum_{y=1}^\infty W_y \approx \sum_{y=y_\text{l}}^{y_\text{u}}W_y
\end{equation}
where $y_\text{l}<y_{\text{max}}<y_\text{u}$ and $y_{\text{max}}$ is the value of $y$ which maximises $W_y$. \cite{dunn2005series} used Stirling's approximation to find that
\begin{equation}
  y_{\text{max}} \approx \frac{x^{2-p}}{\phi(2-p)}
\end{equation}
by treating $W_y$ as a continuous and differentiable function of $y$. The derivation is shown in Appendix \ref{chapter:appendix_compoundPoissonSeries}. Because values of $y$ are positive integers, it would be appropriate to round $y_\text{max}$ accordingly
\begin{equation}
  y_{\text{max}} = \text{max}\left[
    1,\text{round}\left(\frac{x^{2-p}}{\phi(2-p)}\right)
  \right]
  \ .
\end{equation}

$y_\text{l}$ and $y_\text{u}$ can be chosen such that $W_{y_\text{l}}$ and $W_{y_\text{u}}$ are less than $\epsilon W_{y_\text{max}}$ where $\epsilon$ is some small constant, for example, $\epsilon=\euler^{-37}$ will be better than machine precision in 64 bits \citep{dunn2005series}. To prevent overflow problems, it is advised to calculate each term in the summation in log scale \citep{dunn2005series} by using the equation
\begin{equation}
  \ln\left[
    \sum_{y=y_\text{l}}^{y_\text{u}}W_y
  \right]
  = 
  \ln\left(
    W_{y_\text{max}}
  \right)
  +\ln\sum_{y=y_\text{l}}^{y_\text{u}}
  \exp\left[
    \ln\left(W_y\right)-\ln\left(W_{y_\text{max}}\right)
  \right]
  \ .
\end{equation}

@@

Simulations of a compound Poisson-gamma random variable were conducted with the aim to compare how well these density evaluation methods performed. This was done by comparing the evaluated densities using the histogram of simulations and a Q-Q plot. The following compound Poisson-gamma distributions were used in the simulations to capture the variety in the compound Poisson-gamma family: $\CPoisson(1,1,1)$, $\CPoisson(1,100,1)$, $\CPoisson(10,1,1)$, $\CPoisson(100,100,1)$. For each of these distribution, 1\,000 random variables were simulated. Varying $\beta$ is not interesting as this only scales the random variable.

There were a few technical problems with the histogram because the compound Poisson has probability mass at zero and probability density for positive numbers. To correctly represent the empirical density of a compound Poisson random variable, a bar chart was used to show the frequency of a zero and a histogram to show the frequency density of positive numbers. However, the Normal approximation and the saddlepoint approximation does not have mass support at zero. Therefore, to compare these approximate densities to the empirical distribution fairly, a histogram containing both zero and positive samples was used when appropriate.

The evaluation of the p.d.f.~using the saddlepoint approximate required a bit of caution to avoid over/underflow problems. Suppose realisations of $X$ were simulated $\left\{x_1,x_2,x_3,\dotsc,x_n\right\}$. The saddlepoint approximation was computed up to a constant using
\begin{equation}
  p_X(x) \propto
  \exp\left[
    -\frac{\alpha+2}{2(\alpha+1)}
    \ln(x)
    -x\beta
    +\left(
    \frac{x\beta}{\alpha}
    \right)^{\frac{\alpha}{\alpha+1}}\lambda^{\frac{1}{\alpha+1}}(\alpha+1) - k
  \right]
\end{equation}
for 10\,000 equally spaced points from and including the minimum non-zero simulated value to the maximum simulated value. $k$ is some constant which was chosen to be
\begin{equation}
  k =
  \max_{i\in\left\{1,2,3,\dotsc,n\right\}}
  \left[
    -\frac{\alpha+2}{2(\alpha+1)}
    \ln(x_i)
    -x_i\beta
    +\left(
      \frac{x_i\beta}{\alpha}
    \right)^{\frac{\alpha}{\alpha+1}}\lambda^{\frac{1}{\alpha+1}}(\alpha+1)
  \right]
  \ .
\end{equation}
The density was then was normalised by numerically integrating it using the trapezium rule using the 10\,000 evaluated points.

A Q-Q plot is a plot which compares the empirical quantiles with the theoretical quantiles. Let
\begin{equation}
  F_X(x) = \prob(X\leqslant x)
\end{equation}
and
\begin{equation}
  \widehat{F}_X(x) = \frac{1}{n}\cdot\text{max}
  \left[
    \left(\sum_{i=1}^n\mathbb{I}(x_i\leqslant x)\right)-0.5,0
  \right]
  \ ,
\end{equation}
then a Q-Q plot is a parametric plot which plots $\widehat{F}_X^{-1}(p)$ against $F_X^{-1}(p)$ for $p=\frac{0.5}{n},\frac{1.5}{n},\frac{2.5}{n},\dotsc,\frac{n-0.5}{n}$. If $F_X(x)$ and $\widehat{F}_X(x)$ are similar, a Q-Q plot should be a straight line with gradient 1, intercepting the origin. For the exact method and the saddlepoint approximation, $F_X(x)$ was found numerically by evaluating the p.d.f.~at $10\,000$ equally spaced points from and including the minimum to the maximum of the simulated samples, and then summing the required trapeziums. For the saddlepoint approximation, it does not support zero, thus the numerical integration started at a non-zero value. $\widehat{F}_X^{-1}(p)$ was then calculated by interpolation.
